---
title: "PCAWG_all_spectra"
author: "Xinyi"
date: "2023-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# prepare packages 
```{r}
# dataset 
library(PCAWG7) 
# PCA plotting tool  
library(factoextra) 

library(tibble)
library(tidyr)

# heatmap 
library(gplots)

library(ggfortify)
# for dataframe manipulation
library(dplyr)
# for tsne computation
library(Rtsne)
# for umap computation
library(umap)


# General plotting tool
library(ggplot2)
#
library(plotly) 
# for 3D pca plot
library(rgl)


# for tsne computation
library(Rtsne)
# for umap computation
library(umap)

#library(rsvd)
```


```{r}
all_exposure<- as.data.frame(t(PCAWG7::exposure$PCAWG$SBS96))

aa <- PCAWG7::exposure.stats$PCAWG$SBS96
bb <- lapply(aa, rownames)

enframe(bb) %>%
    unnest(value)

data.frame(lapply(bb, "length<-", max(lengths(bb))))
```


# Considering all the mutations types: #SBS96, SBS192, SBS1536, Indel, DBS
```{r}
all_spectra <- as.data.frame(t(PCAWG7::spectra$PCAWG$SBS96
                                     #PCAWG7::spectra$TCGA$SBS96)))
                                     ))
                                     #PCAWG7::spectra$PCAWG$SBS192,
                                     #PCAWG7::spectra$PCAWG$SBS1536,
                                     #PCAWG7::spectra$PCAWG$ID,
                                     #PCAWG7::spectra$PCAWG$DBS78)))
all_spectra <- rownames_to_column(all_spectra,"cancer_type")
# extract cancer type from sample name
for (i in 1:length(all_spectra$cancer_type)){
  tmp <- strsplit(all_spectra$cancer_type[i],"::")[[1]][1]
  all_spectra$cancer_type[i] <- tmp
}
dim(all_spectra)
```
```{r}
all_spectra_subtypes <- all_spectra
```

```{r}
# number of cancer types in the study
count(unique(all_spectra['cancer_type']))
```

```{r}
# combine subtypes to bigger type groups
for (i in 1:length(all_spectra$cancer_type)){
  tmp <- strsplit(all_spectra$cancer_type[i],"-")[[1]][1]
  all_spectra$cancer_type[i] <- tmp
}
```

```{r}
# number of cancer subtypes in the study
count(unique(all_spectra['cancer_type']))
```

# plot variation in LSC
```{r}
PlotSpectraAsSigsWithUncertainty <- function(xx,mean_xx, title = "Mean.as.signature") {
  arrow.tops <- apply(xx, 2, max)
  arrow.bottoms <- apply(xx, 2, min)
  
  options(digits = 2)
  
  bp <- ICAMS::PlotCatalog(
    catalog = mean_xx,
    upper   = TRUE,
    xlabels = TRUE,
    cex     = 0.8,
    ylim    = c(min(arrow.bottoms), max(arrow.tops) + 0.005)
  )
  add.arrows(bp$plot.object, arrow.tops, arrow.bottoms)
  #xx$arrow.tops    <- arrow.tops
  #xx$arrow.bottoms <- arrow.bottoms
  return(invisible(xx))
}
```

```{r}
add.arrows <- function(bp, tops, bottoms) {
  oldopt <- getOption("warn")
  on.exit(options(warn = oldopt))
  options(warn = -1) # Does not seem to turn off warnings
  which0 <- which((tops - bottoms) == 0)
  tops[which0] <- tops[which0] + 1e-5 
  suppressWarnings(
    # Necessary because generates warnings for 0-length arrows
    arrows(
      x0     = bp,
      y0     = tops,    # location of up arrow tips
      y1     = bottoms, # location of down arrow tips
      angle  = 90,      # use "T" arrow heads
      code   = 3,       # use arrow heads at both ends
      length = 0.025    # width of arrow heads
    ))
}
```

```{r, fig.width=10, fig.height=3}
this_spectra <- all_spectra[(all_spectra$cancer_type =="Liver"),-1]
avg.sig <- as.matrix(colMeans(this_spectra[,]))
colnames(avg.sig) <- "Liver"
par(oma = c(1.5,3,1.5,1.5))
#ICAMS::PlotCatalog(avg.sig)
PlotSpectraAsSigsWithUncertainty(this_spectra,avg.sig, title = "Mean.as.signature") 
```

```{r, fig.width=10, fig.height=3}
this_spectra <- all_spectra[(all_spectra$cancer_type =="Lung"),-1]
avg.sig <- as.matrix(colMeans(this_spectra[,]))
colnames(avg.sig) <- "Lung"
par(oma = c(1.5,3,1.5,1.5))
#ICAMS::PlotCatalog(avg.sig)
PlotSpectraAsSigsWithUncertainty(this_spectra,avg.sig, title = "Mean.as.signature") 
```


```{r, fig.width=10,fig.height=3}
this_spectra <- all_spectra[(all_spectra$cancer_type =="Skin"),-1]
avg.sig <- as.matrix(colMeans(this_spectra[,]))
colnames(avg.sig) <- "Skin"
par(oma = c(1.5,3,1.5,1.5))
#ICAMS::PlotCatalog(avg.sig)
PlotSpectraAsSigsWithUncertainty(this_spectra,avg.sig, title = "Mean.as.signature") 
```

```{r, fig.width=10, fig.height=3}
this_spectra <- all_spectra[(all_spectra$cancer_type =="ColoRect"),-1]
avg.sig <- as.matrix(colMeans(this_spectra[,]))
colnames(avg.sig) <- "ColoRect"
par(oma = c(1.5,3,1.5,1.5))
#ICAMS::PlotCatalog(avg.sig)
PlotSpectraAsSigsWithUncertainty(this_spectra,avg.sig, title = "Mean.as.signature") 
```


# Plot Avg count signature 
```{r, fig.height=4, fig.width=15}
this_spectra <- all_spectra[(all_spectra$cancer_type =="Lung"),-1]
avg.sig <- as.matrix(colMeans(this_spectra[,]))
colnames(avg.sig) <- "Lung"
par(oma = c(1.5,3,1.5,1.5))
ICAMS::PlotCatalog(avg.sig)
```

```{r, fig.height=4, fig.width=15}
this_spectra <- all_spectra[(all_spectra$cancer_type =="Skin"),]
avg.sig <- as.matrix(colMeans(this_spectra[,-1]))
colnames(avg.sig) <- "Skin"
par(oma = c(1.5,3,1.5,1.5))
ICAMS::PlotCatalog(avg.sig)
```

```{r, fig.height=4, fig.width=15}
this_spectra <- all_spectra[(all_spectra$cancer_type =="ColoRect"),]
avg.sig <- as.matrix(colMeans(this_spectra[,-1]))
colnames(avg.sig) <- "ColoRect"
par(oma = c(1.5,3,1.5,1.5))
ICAMS::PlotCatalog(avg.sig)
```


```{r}
pca_res <- prcomp(all_spectra[,-1], center = TRUE, scale. = TRUE)
autoplot(pca_res,data = all_spectra[,], colour = 'cancer_type') 
```
```{r}
fviz_pca_ind(pca_res,
             #addEllipses=TRUE, ellipse.level=0.95,
             habillage=all_spectra[, "cancer_type"])
```

```{r}
sc_all_spectra <- NULL
for (i in 0:5)
  sc_all_spectra <- cbind(sc_all_spectra,as.matrix(rowMeans(all_spectra[,(2+i*16):(17+i*16)])))
sc_all_spectra <- as.data.frame(sc_all_spectra)
sc_all_spectra <- cbind(all_spectra[,1], sc_all_spectra)
colnames(sc_all_spectra) <-  c("cancer_type","C>A","C>G","C>T","T>A","T>C","T>G")
sc_all_spectra <- as.data.frame(sc_all_spectra, stringsAsFactors = FALSE)
sc_all_spectra_split <- split(sc_all_spectra,sc_all_spectra$cancer_type)
sc_all_spectra_split <- lapply(sc_all_spectra_split, function(x){x[,-1, drop=F]})
colmean_split <- lapply(sc_all_spectra_split, colMeans)
sum_table <- do.call(cbind,colmean_split)
```

```{r}
View(as.data.frame(sort(colSums(sum_table))))
```

```{r}
#write_csv(as.data.frame(sum_table), "sum_table.csv")
```

```{r}
gplots::heatmap.2(#x = t(sum_table[,-c(7, 12,18, 22)]),
                  x=sum_table,
                  scale = "column",
                  #scale ="row",
                  dendrogram = "none",
                  col = "heat.colors",
                  key = T,
                  keysize = 2,
                  margins = c(5, 5),
                  cex.axis = 1,
                  symm = FALSE,
                  trace = "none")
```

```{r}
pca_res <- prcomp(all_spectra[-c(668,699, 2735),-1], center = TRUE, scale. = TRUE)
autoplot(pca_res,data = all_spectra[-c(668,699, 2735),], colour = 'cancer_type') 
```

```{r}
my_data <- all_spectra[-c(668,699, 2735),]
#write.csv(my_data, file = "my_data.csv")
```

```{r}
# plotting with cancer type and sample ID annotated
fviz_pca_ind(pca_res,
             #addEllipses=TRUE, ellipse.level=0.95,
             habillage=all_spectra[-c(668,699,2735), "cancer_type"])
```

# PLotting PC scatter plot in 3D (first 3 PCs)
## PLot 3d method 1
```{r}
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
```

```{r}
components <- pca_res[["x"]]
components <- data.frame(components)
components$PC2 <- -components$PC2
components$PC3 <- -components$PC3
components = cbind(components, all_spectra[-c(668,699, 2735),1])

tot_explained_variance_ratio <- summary(pca_res)[["importance"]]['Proportion of Variance',]
tot_explained_variance_ratio <- 100 * sum(tot_explained_variance_ratio)


fig <- plot_ly(components, x = ~PC1, y = ~PC2, z = ~PC3, color = ~all_spectra[-c(668,699, 2735),1], colors = gg_color_hue(22)) %>%
  add_markers(size = 12)


fig <- fig %>%
  layout(
    title = "3D plot after PCA"
    #scene = list(bgcolor = "#e5ecf6")
)

fig
```

## plot 3d PCA plot method 2 (Abandoned because the visualization is not very nice)
```{r}
scores = as.data.frame(pca_res$x)
```
```{r}
plot3d(scores[,1:3], 
       size=8,
       #col = seq(nrow(scores))
       col = as.numeric(factor(all_spectra$cancer_type)))
 
# text3d(scores[,1:3],
#        texts=c(rownames(scores)), 
#        cex= 0.7, pos=3)
```

# Importance of each dimension (Variance explained by each PC)
```{r}
fviz_eig(pca_res, addlabels = TRUE)
```
# Some general exploration
## plot point quality
```{r}
fviz_pca_ind(pca_res,
             axes = c(1,2),
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = FALSE     # Avoid text overlapping
             )
```

## plot point/sample contribution 
```{r}
fviz_pca_ind(pca_res,
             axes = c(1,2),
             col.ind = "contrib", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = FALSE     # Avoid text overlapping
             )
```

```{r}
# plotting the samples with top 100 contribution
fviz_pca_ind(pca_res, select.ind = list(contrib = 100))
```

## plot ellipse for the cancer types
```{r}
fviz_pca_ind(pca_res,
             addEllipses=TRUE, ellipse.level=0.95,
             habillage=all_spectra[-c(668,699,2735), "cancer_type"],
             geom = "point")
```

# The Contribution of variables to PCs
## Exploring the variable contributions 
```{r}
fviz_pca_var(pca_res, select.var = list(contrib = 10), repel = TRUE)
```

```{r}
# plot the top one contributing variable
fviz_pca_biplot(pca_res, label="var",
               select.var = list(contrib = 1),
               habillage=all_spectra[-c(668,699,2735), "cancer_type"],
               repel = TRUE)
```

## Contribution to PC1
```{r}
fviz_contrib(pca_res, choice = "var", axes = 1, top = 96) 
# + theme(axis.text = element_text(size = 4.5))
# top 1000 drop to 0.05% contribution 
```

```{r}
test <- fviz_contrib(pca_res, choice = "var", axes = 1)
ts.names <- rownames(test$data)
ts <- test$data[,"contrib"]
table(ts>100/96)
```
```{r}
sorted_index <- order(-ts)

bp <- barplot(ts, 
              #border = NA,
              width = 2,
        col = c(rep("blue",16), 
                rep("yellow",16),
                rep("red",16), 
                rep("grey",16),
                rep("green",16), 
                rep("pink",16)),
        las = 2,
        names.arg = names(ts),
        space=0,
        cex.axis = 1,
        # Adjust when necessary
        ylim= c(0,2.5),
        ylab = "Variable Contribution to PC1")
top_n <- 10
# Adjust when necessary
text(bp[sorted_index[1:top_n]], ts[sorted_index[1:top_n]]+0.15, labels=ts.names[sorted_index[1:top_n]], srt=90, adj = c(0.5,1),xpd = TRUE, pos=3, cex =0.8)
# abline(h = 0.15, col = "#69b3a2", lty=3)

#mtext(names(pca_res$rotation[,2])[1:6], side = 1, line =2, srt=90, cex=1)
legend("top", legend = c("C>A","C>G","C>T","T>A","T>C","T>G"), col = c("blue","yellow","red","grey","green","pink"), 
       bty = "n", pch=20 , pt.cex = 2, cex = 0.8, horiz = T, inset = c(0.05, 0.0005))

```

## Contribution to PC2
```{r}
fviz_contrib(pca_res, choice = "var", axes = 2, top = 10)
```

```{r}
test <- fviz_contrib(pca_res, choice = "var", axes = 2)
ts.names <- rownames(test$data)
ts <- test$data[,"contrib"]
table(ts>100/96)
```
```{r}

sorted_index <- order(-ts)
bp <- barplot(ts, 
              #border = NA,
              width = 2,
        col = c(rep("blue",16), 
                rep("yellow",16),
                rep("red",16), 
                rep("grey",16),
                rep("green",16), 
                rep("pink",16)),
        las = 2,
        names.arg = names(ts),
        space=0,
        cex.axis = 1,
        # Adjust when necessary
        ylim= c(0,4),
        ylab = "Variable Contribution to PC1")
top_n <- 10
# Adjust when necessary
text(bp[sorted_index[1:top_n]], ts[sorted_index[1:top_n]]+0.15, labels=ts.names[sorted_index[1:top_n]], srt=90, adj = c(0.5,1),xpd = TRUE, pos=3, cex =0.8)
# abline(h = 0.15, col = "#69b3a2", lty=3)

#mtext(names(pca_res$rotation[,2])[1:6], side = 1, line =2, srt=90, cex=1)
legend("top", legend = c("C>A","C>G","C>T","T>A","T>C","T>G"), col = c("blue","yellow","red","grey","green","pink"), 
       bty = "n", pch=20 , pt.cex = 2, cex = 0.8, horiz = T, inset = c(0.05, 0.0005))

```

# Loadings of the PCs
## Loadings for PC1
```{r,fig.width=8}
pc_load <- pca_res$rotation[,1]
sorted_index <- order(-abs(pca_res$rotation[,1]))
bp <- barplot(pca_res$rotation[,1], col = c(rep("blue",16), 
                                      rep("yellow",16),
                                      rep("red",16), 
                                      rep("grey",16),
                                      rep("green",16), 
                                      rep("pink",16)),
        las = 2,
        names.arg = names(pca_res$rotation[,1]),
        cex.axis = 1,
        ylim = c(-0.15, 0.03),
        space = 0,
        # width = 8,
        ylab = "PC1 Loadings")
legend("top", legend = c("C>A","C>G","C>T","T>A","T>C","T>G"), col = c("blue","yellow","red","grey","green","pink"), 
   bty = "n", pch=20 , pt.cex = 2, cex = 0.8, horiz = TRUE, inset = c(0.05, 0.05))
top_n <- 5
text(bp[sorted_index[1:top_n]], pc_load[sorted_index[1:top_n]]-0.015, labels=names(pc_load[sorted_index[1:top_n]]), srt=90, 
      #adj = c(0.5,1),
    xpd = T, pos=3, cex =0.8)
#abline(h = -0.12, col = "#69b3a2", lty=3)
#text(bp[32], y = -1.2, labels = "y = -1.2", col= "#69b3a2")

# legend("top", legend = c("C>A","C>G","C>T","T>A","T>C","T>G"), col = c("blue","yellow","red","grey","green","pink"), horiz = TRUE, x.intersp = 1.5, y.intersp = 1.5, inset = c(0, -0.1), y = 1.1)
```

## Loadings for PC2
```{r, fig.width=8}
pc_load <- pca_res$rotation[,2]
sorted_index <- order(-abs(pca_res$rotation[,2]))

bp <- barplot(pc_load, 
              #border = NA,
              width = 2,
        col = c(rep("blue",16), 
                rep("yellow",16),
                rep("red",16), 
                rep("grey",16),
                rep("green",16), 
                rep("pink",16)),
        las = 2,
        names.arg = names(pca_res$rotation[,2]),
        space=0,
        cex.axis = 1,
        ylim= c(-0.2,0.2),
        ylab = "PC2 Loadings")
top_n <- 5
text(bp[sorted_index[1:top_n]], 
     #pc_load[sorted_index[1:top_n]]+0.015, 
     ifelse(pc_load[sorted_index[1:top_n]]>0,pc_load[sorted_index[1:top_n]]+0.015,pc_load[sorted_index[1:top_n]]-0.025),
     labels=names(pc_load[sorted_index[1:top_n]]), srt=90, adj = c(0.5,1),xpd = TRUE, pos=3, cex =0.8)
 abline(h = 0.15, col = "#69b3a2", lty=3)
 abline(h = -0.15, col = "#69b3a2", lty = 3)
#mtext(names(pca_res$rotation[,2])[1:6], side = 1, line =2, srt=90, cex=1)
legend("topright", legend = c("C>A","C>G","C>T","T>A","T>C","T>G"), col = c("blue","yellow","red","grey","green","pink"), 
       bty = "n", pch=20 , pt.cex = 2, cex = 0.8, horiz = FALSE, inset = c(0.05, 0.12), yjust = 1.2)

```
## Loadings for PC3
```{r, fig.width=8}
pc_load <- pca_res$rotation[,3]
sorted_index <- order(-abs(pca_res$rotation[,3]))

bp <- barplot(pc_load, 
              #border = NA,
              width = 2,
        col = c(rep("blue",16), 
                rep("yellow",16),
                rep("red",16), 
                rep("grey",16),
                rep("green",16), 
                rep("pink",16)),
        las = 2,
        names.arg = names(pca_res$rotation[,2]),
        space=0,
        cex.axis = 1,
        ylim= c(-0.2,0.35),
        ylab = "PC3 Loadings")
top_n <- 5
text(bp[sorted_index[1:top_n]], 
     #pc_load[sorted_index[1:top_n]]+0.015, 
     ifelse(pc_load[sorted_index[1:top_n]]>0,pc_load[sorted_index[1:top_n]]+0.015,pc_load[sorted_index[1:top_n]]-0.025),
     labels=names(pc_load[sorted_index[1:top_n]]), srt=90, adj = c(0.5,1),xpd = TRUE, pos=3, cex =0.8)
 #abline(h = 0.15, col = "#69b3a2", lty=3)
 #abline(h = -0.15, col = "#69b3a2", lty = 3)
#mtext(names(pca_res$rotation[,2])[1:6], side = 1, line =2, srt=90, cex=1)
legend("topright", legend = c("C>A","C>G","C>T","T>A","T>C","T>G"), col = c("blue","yellow","red","grey","green","pink"), 
       bty = "n", pch=20 , pt.cex = 2, cex = 0.8, horiz = FALSE, inset = c(0.05, 0.12), yjust = 1.2)

```
# Comparison with tsne
```{r}
tsne_res <- Rtsne(all_spectra[-c(668,699),-1])

# Conversion of matrix to dataframe
tsne_plot <- data.frame(x = tsne_res$Y[,1],
                        y = tsne_res$Y[,2], colour = all_spectra[-c(668,699),"cancer_type"])
 
# Plotting the plot using ggplot() function
#ggplot2::ggplot(tsne_plot,label=Species)+ geom_point(aes(x=x,y=y))
ggplot(tsne_plot, aes(x, y, colour = colour)) +
  geom_point()
```

# Comparison with Umap
```{r}
umap_res = umap(all_spectra[-c(668,699),-1], n_components = 2, k = 10) 
layout <- umap_res[["layout"]] 
layout <- data.frame(layout) 
final <- cbind(layout, all_spectra[-c(668,699),1]) 
colnames(final) <- c('X1', 'X2', 'cancer_type') 

# fig <- plot_ly(final, x = ~X1, y = ~X2, split = ~cancer_type,  type = 'scatter', mode = 'markers')%>%  
#   layout(  
#     plot_bgcolor = "#e5ecf6",
#     legend=list(title=list(text='cancer_type')), 
#     xaxis = list( 
#       title = "0"),  
#     yaxis = list( 
#       title = "1")) 
# fig

final %>%
  ggplot(aes(x = X1, 
             y = X2, 
             color = cancer_type))+
  geom_point()+
  labs(x = "UMAP1",
       y = "UMAP2",
      subtitle = "UMAP plot")
```


